One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2019-10-10 paper.tex
more waffle cut

As an alternative to the quantile maps we use excursion sets and excursion functions \citep{bolin_excursion_2015}.  This is a recent addition to the literature so we take some time here to define and elaborate on the concepts used.  This approach to representing variability in stochastic processes specifically considers the joint probability of events across multiple locations.  The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e. the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$. For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
% Somehow keep this information somewhere in the paper:
%This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.  Three such realizations are in \autoref{fig:intensity-realizations}.  There are some notable differences when compared to the mean and quantile summary maps.  Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.  In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.  The summary plots hide the spatial structure the model has been able to estimate from the data.  In general, the mean of a random field will contain less spatial structure than realisations of the field.
%
%\begin{figure}
%	\includegraphics[scale=0.35]{figures/intensity_realized.png}
%	\caption{Three realizations of the posterior intensity field}
%	\label{fig:intensity-realizations}
%\end{figure} 
\begin{equation*}
E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
\end{equation*}
Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.  As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.  Negative excursion sets can be similarly defined.  Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.  An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.

@@

%A key feature of these plots is that they display either discrete sub-regions of the area of interest (for excursion sets) or the probability of membership of such discrete sub-regions (for excursion functions).  We believe the risk of interpreting these maps as intensities is much lower than for the quantile plots.  The fact that it is possible to estimate discrete regions may also be useful to communicate results to conservation practitioners, land and resource managers and policy decision-makers.  We anticipate the excursion function will be most useful in communicating to audiences with statistical training. % [RJC - I think statistically savvy audiences will readily grasp the excursion functions. Do you mean "without" statistical training?]

Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.  We describe here how to estimate the posterior for the abundance. Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, the \textit{expected} abundance.  We approximate the posterior $\pi(N | \lambda)$ by a monte-carlo method.  Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$. Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$. \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.  For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance. We note here that in a GAM approach it is common to only report the expected abundance and the error in expected abundance.  This does not include the additional variability that comes from the Poisson distribution with a rate parameter equal to the expected abundance.

Finally, we show that the method has simultaneusly identified a plausible detection function.  \autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.  We note again that all model summaries shown in this section are based on monte-carlo sampling of the joint posterior of all model parameters.  As such all the plots incorporate the uncertainty in the detection process.  
\begin{figure}
	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
	\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
	\label{fig:realized-abundance-posterior}
\end{figure}

\begin{figure}
	\includegraphics[scale=0.6]{figures/realized_abundance_vs_exp.png}
	\caption{Solid:  Posterior of realized abundance.  Dashed:  Poisson PMF with rate parameter $\mathbb{E}[\Lambda]$}
	\label{fig:realized-abundance-comparison}
\end{figure}

\begin{figure}
	\includegraphics[scale=0.6]{figures/halfnormal.png}
	\caption{Posterior detection function}
	\label{fig:half-normal}
\end{figure}

@@

The most common method to analyse distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.  The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.  However, these two approaches are related and an analysis of the \akepa{} data using the GAM approach provides similar point estimates.  There is a correspondence between SPDEs and smoothing splines: the smoothing spline is an estimator of the posterior mean of the SPDE \citep{kimeldorf_spline_2002}.  The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.  
In addition, via the point process model, we have avoided the need to bin data into counts, which may at first appear as though we have avoided some loss of information. This is not the case since we also applied the standard distance sampling assumption of constant density within each point transect.  This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points.  The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs at the level of approximating the GRF.  This is in contrast to motivating the discretisation by binning the observed data which requires consideration of zero-inflation and dispersion.

We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.  As such our approach is extendible to more complicated contexts such as marked point processes.

@@

\subsection*{Possible future directions}

Here we applied the iterated INLA approach in order to estimate the half-normal detection function which is non-linear in its $\log \sigma$ parameter.  We note that this approach provides a general framework for estimating parametric functions from data and is not restricted to detection functions within a distance sampling model.  The \texttt{inlabru} software allows users to specify the model in an intuive and modular way.  Users write individual functions to define (possibly non-linear) model components and specify how these are combined to form the additive predictor.  By additive we mean a predictor that is a sum of model components, and not necessarily additive in terms of parameters to be estimated.

>>>>

As an alternative to the quantile maps we use excursion sets and excursion functions \citep{bolin_excursion_2015}.  This is a recent addition to the literature so we take some time here to define and elaborate on the concepts used.  This approach to representing variability in stochastic processes specifically considers the joint probability of events across multiple locations.  The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e. the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$. For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
% Somehow keep this information somewhere in the paper:
%This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.  Three such realizations are in \autoref{fig:intensity-realizations}.  There are some notable differences when compared to the mean and quantile summary maps.  Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.  In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.  The summary plots hide the spatial structure the model has been able to estimate from the data.  In general, the mean of a random field will contain less spatial structure than realisations of the field.
%

\begin{equation*}
E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
\end{equation*}
Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.  As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.  Negative excursion sets can be similarly defined.  Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.  An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.

@@

Given the issues of interpreting maps that simultaneously show summaries of the posterior random field it can often be informative to look directly at realisations themselves. Three such realizations are in \autoref{fig:intensity-realizations}.  
\begin{figure}
	\includegraphics[scale=0.35]{figures/intensity_realized.png}
	\caption{Three realizations of the posterior intensity field}
	\label{fig:intensity-realizations}
\end{figure}
There are some notable differences when compared to the mean and quantile summary maps.  Note each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.  In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only considered the posterior mean of the field.  The summary maps hide the spatial structure the model has been able to estimate from the data.

Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.  We describe here how to estimate the posterior for the abundance. Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance - the expected abundance.  We approximate the posterior $\pi(N | \lambda)$ by a monte-carlo method.  Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$. Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$. \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.  For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
\begin{figure}
	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
	\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
	\label{fig:realized-abundance-posterior}
\end{figure}
\begin{figure}
	\includegraphics[scale=0.6]{figures/realized_abundance_vs_exp.png}
	\caption{Solid:  Posterior of realized abundance.  Dashed:  Poisson PMF with rate parameter $\mathbb{E}[\Lambda]$}
	\label{fig:realized-abundance-comparison}
\end{figure}
We note here that in a GAM approach it is common to only report the estimated expected abundance and the uncertainty in this estimate.  This does not include the additional variability that comes from the Poisson distribution with a rate parameter equal to the expected abundance.

Finally, we show that the one-stage model has estimated a plausible detection function.  \autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.  We note again that all model summaries shown in this section are based on monte-carlo sampling of the joint posterior of all model parameters.  As such all the plots incorporate the uncertainty in the detection process.  
\begin{figure}
	\includegraphics[scale=0.6]{figures/halfnormal.png}
	\caption{Posterior detection function}
	\label{fig:half-normal}
\end{figure}

@@

The most common method to analyse distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.  The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.  However, these two approaches are related and an analysis of the \akepa{} data using the GAM approach provides similar point estimates.  There is a correspondence between SPDEs and smoothing splines: the smoothing spline is an estimator of the posterior mean of the SPDE \citep{kimeldorf_spline_2002}.  The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.  
In addition, it can appear as though the point process model avoids the need to bin data into counts and so avoids some loss of information. This is not the case since we also applied the standard distance sampling assumption of constant intensity within each point transect.  This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points across the whole study region.  The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs during the step of approximating the GRF.  This is in contrast to motivating the discretisation by the act of binning the data into counts so as to create a response variable that can be analysed using a GAM model.  We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.  As such our approach is extendible to more complicated contexts such as marked point processes.

@@

\subsection*{Possible future research}

Here we applied the iterated INLA approach in order to estimate the half-normal detection function which is non-linear in its $\log \sigma$ parameter.  We note that this approach provides a general framework for estimating parametric functions from data and is not restricted to detection functions in a distance sampling model.  The \texttt{inlabru} software allows users to specify the model in an intuive and modular way.  Users write individual functions to define (possibly non-linear) model components and specify how these are combined to form the additive predictor.  By additive we mean a predictor that is a sum of model components, and not necessarily additive in terms of parameters to be estimated.
